Towards Correct Legion

### 2017-08-25 correctnesshpc/paper.tex
Small tweak to LANL org (lab vs. labs).
add citations

\title{Towards Correct Legion}
\author{George Stelle}
\affiliation{\institution{Los Alamos National Laboratories}}
\email{stelleg@lanl.gov}

@@

\author{Pat McCormick}
\affiliation{\institution{Los Alamos National Laboratories}}
\email{pat@lanl.gov}

@@

The strongest guarantees of correctness come from the area of formal
specifications and proofs. A specification is a description of how a program
should behave, while proofs are used to show that an implementation conforms to
a specification. \emph{Formal} specifications and proofs are defined as
specifications and proofs that have been embedded in a machine-checked formal
logic \cite{Coq, isabelle...}. This process of formalizing specifications and
proofs in machine checked logics has been shown to catch common errors in
previously informal proofs \cite{papers}, and therefore is now considered
to provide the highest assurance of correctness currently possible \cite{?}.

In addition, recent work has shown that modular, composable formal
specifications enable formal correctness proofs for large software systems that
were previously unattainable \cite{?}. By integrating a formal specification
tightly with an implementation, one can achieve the highest levels of
confidence in code currently possible \cite{?}. For example, the Compcert
project implements an optimizing C compiler and proves it correct. The success
of Compcert in removing compiler bugs has been well documented. We will discuss
later how Compcert could be used in building verifiable HPC systems.

In HPC, programming models have become an increasingly important way to ensure
that code stays maintainable, performant, and portable \cite{?}. We claim that
these programming models provide a good target for formalization: any
correctness arguments can be leveraged by every program using the model. It is
exactly this kind of modularization of proofs that has been so successful in
proving programs correct \cite{?}.

@@

Legion is a programming model for high performance computing that enables
reasoning about locality and concurrency via logical regions. It has the
attractive property that it is guaranteed to preserve serial semantics for
memory operations within logical regions. This allows programmers to avoid the
burden of reasoning about the non-determinism that arises from concurrent
computation. For this paper, we'll be formalizating work on from Triecher et.
al \cite{ooplsa13}. While there has been much work extending Legion since then,
the core semantics has stayed relatively stable \cite{regent, deppart}.

By partitioning regions into \emph{subregions}, and tracking \emph{privileges}, 
Legion is able to ensure that certain code won't interfere or alias, and can
therefore be safely, at least with respect to those memory regions, run in
parallel. This theory has been put into practice to impressive effect, and has
been shown to scale both within a node and between \cite{legionpapers}. That
said, there are still challenges --subtle bugs that occasionally show up in the
implementation. It is possible that these kinds of bugs could manifest
themselves in ways that would lead to incorrect conclusions when informing
major decisions, something we'd generally like to avoid. One goal of this paper
is to move towards formally and systematically removing \emph{all} bugs from
the programming model, giving everyone who uses is far more certainty that 
their code is correct.

@@

Coq, in particular, has proven itself to be a particularly potent dependently
typed proof assistant. At its core is the Calculus of Inductive Constructions,
a small dependently typed programming language and logic, capable of powerful
abstractions for formal proofs \cite{cic}. In the last 10 years great strides
have been made in using it to both build systems and prove useful properties
about them \cite{compcert, certicoq, chlipala...}.

Specifications come in a wide range of formats. Many, like the OpenMP
specification, are very informal, primarily consisting of english descriptions
of how different constructs should behave \cite{openmp}. Others, like Legion's,
are written by hand in the using standard notation for logical relations, etc.
For this paper, when we refer to a \emph{formal} specification, we are
referring to a specification defined in a formal logic, eg. Coq.

@@

A program without a specification cannot be incorrect, it can only be
surprising \footnote{Paraphrase of J.J. Horning, 1982}. One of the most
important and challenging aspects of proving correctness is defining a formal
specification of how a program should behave. Indeed, there is currently a
large NSF effort involving many of the best minds in programming languages
dedicated to improving our ability to do just this \cite{deepspec}. The choice
of specification has far-reaching consequences in how easy it is to later prove
correctness properties.

One thing that makes Legion unique among high performance runtime systems is
that it provides a specification of its operational semantics \cite{oopsla13}.
We take this specification of and formalize it in Coq, enabling us to prove
lemmas and theorems about it. See Figure~\ref{syntax} for the Coq
implementation of the syntax, and Figure~\ref{semantics}and semantics. All code
is available online at \texttt{https://github.com/lanl/formallegion}.

@@

All of these issues are addressable given enough time and effort. Some offer
a nice opportunity for modularity within the tasking formalization. For example, 
ordering tasks based on dependencies of values is common accross all tasking
systems \cite{cilk, hpx}, and is likely a sort of sub-semantics that they could
all share, reducing proof burden across different run-times.

@@

This is one oft-overlooked advantage to formal reasoning: it allows reasoning
about \emph{any} property of a system, not just correctness. For example, there
is the work by \cite{?} working towards formally verified complexity of code.
Chlipala et. al. have also used techniques like those described here to construct
and verify hardware description, where performance must be reasoned about down
to single cycles \cite{chlipala}.  This is one area where other tasking
systems, like Cilk, have a better story than Legion. Blumofe et al. informally
describe bounds on runtime overheads, which has been realized in practice as
low overheads in the actual runtime \cite{cilk}.

@@

While we've focused on Legion, there are other parallel programming systems
that offer exciting opportunities for formalization. While some (looking at
you, OpenMP), are likely much to large and complex to formalize without
herculean effort, others have simple enough semantics that formalization is
likely within reach. One in particular that is attractive due to its simplicity
is the recent work on parallel extensions to LLVM, the Tapir project
\cite{tapir}. Building on existing formalization efforts for LLVM
\cite{verillvm}, extending existing semantics with the Tapir instructions for
concurrency should be viable. Again, there would be many challenges,
particularly in the area of formal reasoning about memory models, but the
approach should be viable.

>>>>

\title{Towards Correct Legion}
\author{George Stelle}
\affiliation{\institution{Los Alamos National Laboratory}}
\email{stelleg@lanl.gov}

@@

\author{Pat McCormick}
\affiliation{\institution{Los Alamos National Laboratory}}
\email{pat@lanl.gov}

@@

The strongest guarantees of correctness come from the area of formal
specifications and proofs. A specification is a description of how a program
should behave, while proofs are used to show that an implementation conforms to
a specification. \emph{Formal} specifications and proofs are defined as
specifications and proofs that have been embedded in a machine-checked formal
logic \cite{bertot2013interactive, nipkow2002isabelle, pfenning1999system, jackson1994nuprl}. This process of formalizing specifications and
proofs in machine checked logics has been shown to catch common errors in
previously informal proofs \cite{clarke1996formal}, and therefore is now considered
to provide the highest assurance of correctness currently possible \cite{?}.

In addition, recent work has shown that modular, composable formal
specifications enable formal correctness proofs for large software systems that
were previously unattainable \cite{weng2016deepspec}. By integrating a formal specification
tightly with an implementation, one can achieve the highest levels of
confidence in code currently possible \cite{weng2016deepspec}. For example, the Compcert
project implements an optimizing C compiler and proves it correct. The success
of Compcert in removing compiler bugs has been well documented. We will discuss
later how Compcert could be used in building verifiable HPC systems.

In HPC, programming models have become an increasingly important way to ensure
that code stays maintainable, performant, and portable \cite{diaz2012survey}. We claim that
these programming models provide a good target for formalization: any
correctness arguments can be leveraged by every program using the model. It is
exactly this kind of modularization of proofs that has been so successful in
proving programs correct \cite{weng2016deepspec}.

@@

Legion is a programming model for high performance computing that enables
reasoning about locality and concurrency via logical regions. It has the
attractive property that it is guaranteed to preserve serial semantics for
memory operations within logical regions. This allows programmers to avoid the
burden of reasoning about the non-determinism that arises from concurrent
computation. For this paper, we'll be formalizating work on from Triecher et.
al \cite{treichler2013language}. While there has been much work extending Legion since then,
the core semantics has stayed relatively stable \cite{slaughter2015regent, deppart}.

By partitioning regions into \emph{subregions}, and tracking \emph{privileges}, 
Legion is able to ensure that certain code won't interfere or alias, and can
therefore be safely, at least with respect to those memory regions, run in
parallel. This theory has been put into practice to impressive effect, and has
been shown to scale both within a node and between \cite{bauer2012legion, treichler2014realm}. That
said, there are still challenges --subtle bugs that occasionally show up in the
implementation. It is possible that these kinds of bugs could manifest
themselves in ways that would lead to incorrect conclusions when informing
major decisions, something we'd generally like to avoid. One goal of this paper
is to move towards formally and systematically removing \emph{all} bugs from
the programming model, giving everyone who uses is far more certainty that 
their code is correct.

@@

Coq, in particular, has proven itself to be a particularly potent dependently
typed proof assistant. At its core is the Calculus of Inductive Constructions,
a small dependently typed programming language and logic, capable of powerful
abstractions for formal proofs \cite{bertot2013interactive}. In the last 10 years great strides
have been made in using it to both build systems and prove useful properties
about them \cite{leroy2012compcert, anand2017certicoq, chlipala2009effective}.

Specifications come in a wide range of formats. Many, like the OpenMP
specification, are very informal, primarily consisting of english descriptions
of how different constructs should behave \cite{dagum1998openmp}. Others, like Legion's,
are written by hand in the using standard notation for logical relations, etc.
For this paper, when we refer to a \emph{formal} specification, we are
referring to a specification defined in a formal logic, eg. Coq.

@@

A program without a specification cannot be incorrect, it can only be
surprising \footnote{Paraphrase of J.J. Horning, 1982}. One of the most
important and challenging aspects of proving correctness is defining a formal
specification of how a program should behave. Indeed, there is currently a
large NSF effort involving many of the best minds in programming languages
dedicated to improving our ability to do just this \cite{weng2016deepspec}. The choice
of specification has far-reaching consequences in how easy it is to later prove
correctness properties.

One thing that makes Legion unique among high performance runtime systems is
that it provides a specification of its operational semantics \cite{treichler2013language}.
We take this specification of and formalize it in Coq, enabling us to prove
lemmas and theorems about it. See Figure~\ref{syntax} for the Coq
implementation of the syntax, and Figure~\ref{semantics}and semantics. All code
is available online at \texttt{https://github.com/lanl/formallegion}.

@@

All of these issues are addressable given enough time and effort. Some offer
a nice opportunity for modularity within the tasking formalization. For example, 
ordering tasks based on dependencies of values is common accross all tasking
systems \cite{blumofe1995cilk, kaiser2014hpx}, and is likely a sort of sub-semantics that they could
all share, reducing proof burden across different run-times.

@@

This is one oft-overlooked advantage to formal reasoning: it allows reasoning
about \emph{any} property of a system, not just correctness. For example, there
is the work by \cite{?} working towards formally verified complexity of code.
Chlipala et. al. have also used techniques like those described here to construct
and verify hardware description, where performance must be reasoned about down
to single cycles \cite{braibant2013formal}.  This is one area where other tasking
systems, like Cilk, have a better story than Legion. Blumofe et al. informally
describe bounds on runtime overheads, which has been realized in practice as
low overheads in the actual runtime \cite{blumofe1995cilk}.

@@

While we've focused on Legion, there are other parallel programming systems
that offer exciting opportunities for formalization. While some (looking at
you, OpenMP), are likely much to large and complex to formalize without
herculean effort, others have simple enough semantics that formalization is
likely within reach. One in particular that is attractive due to its simplicity
is the recent work on parallel extensions to LLVM, the Tapir project
\cite{schardl2017tapir}. Building on existing formalization efforts for LLVM
\cite{zhao2012formalizing}, extending existing semantics with the Tapir instructions for
concurrency should be viable. Again, there would be many challenges,
particularly in the area of formal reasoning about memory models, but the
approach should be viable.
